---
title: "Total spending: Direct, Total, and when each is right"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Total spending: Direct, Total, and when each is right}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Two questions that sound the same but aren't

"Total spending" means two different things depending on whether the question
is about one government or several:

1. **"What did my county spend in total, a decade ago vs today?"** — one
   government, tracked over time. Either `direct` or `total` spending answers
   this correctly, as long as the same concept is used for both years.
2. **"How do all the counties in my state compare, a decade ago vs today,
   against the neighboring state?"** — several governments, summed together.
   Here only `direct` gives the right answer; summing `total` across
   governments double-counts money that passes between them.

`cog_spending()`'s `expenditure_concept` argument (`"direct"` or `"total"`)
controls which of these a query answers. This vignette walks through both
questions with code that actually runs against the package's bundled fixture
corpus, then explains why the second question refuses `"total"` outright.

```{r}
library(uscogdata)

# Point at the bundled offline fixture (years 2011, 2012, 2019, 2020, all 50
# states) so this vignette knits without network access. In real use,
# USCOGDATA_URL is instead set to the published corpus URL -- see README.md.
Sys.setenv(USCOGDATA_URL = paste0(
  system.file("extdata/fixture_corpus", package = "uscogdata"), "/"
))
```

The fixture doesn't carry 2017 or the present year, so the examples below use
the closest years it does ship -- **2012 and 2020** -- in place of "2017 vs
today" / "ten years ago vs today". Point `USCOGDATA_URL` at the published
corpus and swap in real years; the mechanics are identical.

# Archetype 1: one government's own trend

For a single government, `total` is a legitimate way to describe "everything
this government spent, including money it handed to other governments to
spend on its behalf":

```{r}
al_total <- cog_spending(
  "010000226085",                 # Alabama, the state government
  years = c(2012, 2020),
  category = "Highways",
  expenditure_concept = "total"
)
al_total
```

The `intergovernmental` rows are what `"total"` adds on top of `"direct"`
(`capital` + `operations`): Alabama's own payments out to counties and
cities for highway work. Because this query only ever concerns Alabama,
including that piece is safe -- there's no other government's number it
could be double-counted against.

`"direct"` (the default) answers the same trend question just as validly:

```{r}
al_direct <- cog_spending(
  "010000226085", years = c(2012, 2020), category = "Highways"
  # expenditure_concept = "direct" is the default; shown here for contrast
)
al_direct
```

Both are internally consistent series. What breaks the comparison is
**switching concepts between the two years being compared** -- e.g. `direct`
for 2012 and `total` for 2020 -- which manufactures a trend that isn't
really there. Pick one concept for a given question and hold it fixed across
every year in the series.

# Archetype 2: a cross-government rollup

`cog_geographic_rollup()` sums spending across state/county/city layers for
a place. Its default -- and, as shown below, its *only* accepted value for
`expenditure_concept` -- is `"direct"`:

```{r}
fl_rollup <- cog_geographic_rollup(
  govids = list(
    state  = "120000226351",                   # Florida
    county = c("121011212191", "121099101897") # Broward + Palm Beach
  ),
  category = "Highways",
  years = c(2012, 2020)
)
fl_rollup
```

For the neighboring state, the comparison is a single government, so it's a
plain `cog_spending()` call rather than a rollup:

```{r}
ga_state <- cog_spending(
  "130000226087", years = c(2012, 2020), category = "Highways" # Georgia
)
ga_state
```

Now the same rollup, but asking for `expenditure_concept = "total"`:

```{r, error = TRUE}
cog_geographic_rollup(
  govids = list(state = "120000226351", county = "121011212191"),
  category = "Highways",
  years = 2020,
  expenditure_concept = "total"
)
```

`cog_geographic_rollup()` (and `cog_peer_compare()`, for the same reason)
refuses `"total"` outright rather than silently returning an inflated
number. The next section is why.

# The mechanism

Suppose Alabama gives a county $10M toward a highway project. That $10M
shows up **twice** in the underlying corpus:

- Once on Alabama's own record, coded `M44` ("to local governments,
  Highways") -- Alabama's intergovernmental leg.
- Again on the county's record, coded `E44` / `F44` ("Highways, current
  operations" / "capital outlay") -- the county's direct spending, because
  the county is the government that actually lets the contract and pays the
  paving crew.

`direct` (item codes `E`/`F`/`G`) only ever counts the second of those --
the government that actually did the spending. `total` (Direct plus the
`M`/`L` intergovernmental legs) counts the first one *as well*, which is
exactly right for describing Alabama's own budget: Alabama's `total`
genuinely includes the $10M it committed to highways, whether it built the
road itself or paid the county to. But sum `total` across Alabama **and**
the county, and that $10M is counted twice -- once as Alabama's payment out,
once as the county's spending in -- reporting $20M of highway work for $10M
actually spent.

This is exactly the shape of query `cog_geographic_rollup()` exists to run
(summing across layers of government), so it refuses `"total"` rather than
silently overstating every multi-layer figure it produces.

# How big is the risk in practice

Intergovernmental transfers aren't evenly distributed by government type.
Measured on the full published corpus, intergovernmental spending as a
share of a government's own Direct spending is:

| Government type | Intergovernmental / Direct |
|---|---|
| State  | 17.2% |
| County | 1.8%  |
| City   | 0.8%  |

So the Direct/Total choice matters overwhelmingly for **state** governments
-- a state's Total genuinely differs from its Direct by a meaningful margin,
while for a county or city the two are close. That's also why the mistake
this vignette warns about is easy to make unnoticed at the county/city level
and costly at the state level: rolling up every government in a state using
`total` instead of `direct` overstates the true figure -- measured at 7.6%
for Alabama in FY2019, and 11.6% nationally.

# Why Total = Direct + M + L, not Direct + M

It's tempting to assume `total` only needs to add `M` (payments to local
governments). But not all of the money a county or city receives arrives
directly from its state as an `M` payment -- some flows through as `L`
(payments *to* the state government), which the state government then
redistributes as `M`. On the published corpus, `L` is 0 for state
governments (a state has no "payments to the state government" leg of its
own) but is 91.6% the size of `M` for counties and 188.3% the size of `M`
for cities -- so a `total` that omitted `L` would silently undercount Total
specifically for local governments. `cog_spending(expenditure_concept =
"total")` includes both legs (excluding the `L--` family-total rollup row,
which would double-count its own components).

# Composition rules

- `expenditure_concept` (whose spending counts -- Direct vs Direct plus
  intergovernmental) is **orthogonal** to `basis` (which vintage of the
  item-code space a query resolves against -- `"harmonized"` vs `"raw"`).
  They combine freely: `expenditure_concept = "total", basis = "raw"` is a
  valid, meaningful query, and so is every other pairing.
- `expenditure_concept = "total"` is **mutually exclusive** with `recipe`: a
  recipe already defines its own component codes (some recipes have their
  own matching intergovernmental counterpart recipe instead -- see
  `cog_recipes()` and the "firing suggestion" notes surfaced in
  `cog_spending()`'s provenance), so layering a second, generic `total`
  union on top of a recipe query has no well-defined meaning. Passing both
  together aborts with an error naming the conflict.
- `expenditure_concept` is a **spending-only** concept: `cog_revenue()`
  doesn't expose it (revenue's own intergovernmental codes are a different
  axis -- see `?cog_revenue`).

# Summary

- Comparing one government to itself over time: `"direct"` or `"total"`
  both work -- pick one and hold it fixed across every year compared.
- Comparing or summing across governments -- counties within a state, a
  state against its neighbor, cities against counties: use `"direct"`.
  `cog_geographic_rollup()` and `cog_peer_compare()` enforce this by
  refusing `"total"`.
- `"total"` = Direct (`E`/`F`/`G`) + intergovernmental (`M` to local
  governments + `L` to the state government, excluding the `L--`
  family-total row).
