---
title: "Total spending: Primary, Direct, Total, and when each is right"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Total spending: Primary, Direct, Total, and when each is right}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Two questions that sound the same but aren't

"Total spending" means two different things depending on whether the question
is about one government or several:

1. **"What did my county spend in total, a decade ago vs today?"** — one
   government, tracked over time. Any concept answers this correctly, as
   long as the same concept is used for both years.
2. **"How do all the counties in my state compare, a decade ago vs today,
   against the neighboring state?"** — several governments, summed together.
   Here only a non-intergovernmental concept (`primary` or `direct`) gives
   the right answer; summing `total` across governments double-counts money
   that passes between them.

`cog_spending()`'s `expenditure_concept` argument controls which of these a
query answers, via three nested concepts defined as sets of the crosswalk's
`spend_subtype` values (never item-code first letters — the letter `Y` alone
spans revenue, expenditure, and balance codes):

- `"primary"` (the default) — the government's own service provision:
  `operations` + `capital` + `assistance`.
- `"direct"` — Census's published Direct Expenditure: `primary` plus
  `interest` on debt and `insurance_benefits` (e.g. pension payments).
- `"total"` — `direct` plus the `intergovernmental` leg.

This vignette walks through both questions with code that actually runs
against the package's bundled fixture corpus, then explains why the second
question refuses `"total"` outright.

Before any of the numbers below: every amount column here — `amt_nominal`,
`amt_real`, and their `amt_per_capita_*` counterparts — is in **full US
dollars**. The raw Census files report **thousands of dollars** and the
corpus preserves that in its own `amt` column, but the verbs multiply by 1000
on the way out. So `amt_nominal = 1317000` means $1.317 million, not $1.317
billion. Do not scale it again.

```{r}
library(uscogdata)

# Point at the bundled offline fixture (years 2011, 2012, 2019, 2020, all 50
# states) so this vignette knits without network access. In real use,
# USCOGDATA_URL is instead set to the published corpus URL -- see README.md.
Sys.setenv(USCOGDATA_URL = paste0(
  system.file("extdata/fixture_corpus", package = "uscogdata"), "/"
))
```

The fixture doesn't carry 2017 or the present year, so the examples below use
the closest years it does ship -- **2012 and 2020** -- in place of "2017 vs
today" / "ten years ago vs today". Point `USCOGDATA_URL` at the published
corpus and swap in real years; the mechanics are identical.

# Archetype 1: one government's own trend

For a single government, `total` is a legitimate way to describe "everything
this government spent, including money it handed to other governments to
spend on its behalf":

```{r}
al_total <- cog_spending(
  "010000226085",                 # Alabama, the state government
  years = c(2012, 2020),
  category = "Highways",
  expenditure_concept = "total"
)
al_total
```

The `intergovernmental` rows are what `"total"` adds on top of the
non-intergovernmental subtypes (here `capital` + `operations`): Alabama's
own payments out to counties and cities for highway work. Because this
query only ever concerns Alabama, including that piece is safe -- there's
no other government's number it could be double-counted against.

`"primary"` (the default) answers the same trend question just as validly
(for Highways, which maps only to operations/capital codes, `"primary"` and
`"direct"` coincide -- there is no highway-specific interest or insurance
benefit to add):

```{r}
al_primary <- cog_spending(
  "010000226085", years = c(2012, 2020), category = "Highways"
  # expenditure_concept = "primary" is the default; shown here for contrast
)
al_primary
```

Both are internally consistent series. What breaks the comparison is
**switching concepts between the two years being compared** -- e.g. `direct`
for 2012 and `total` for 2020 -- which manufactures a trend that isn't
really there. Pick one concept for a given question and hold it fixed across
every year in the series.

# Archetype 2: a cross-government rollup

`cog_geographic_rollup()` sums spending across state/county/city layers for
a place. Its default is `"primary"`, and (as shown below) it accepts only
the non-intergovernmental concepts, `"primary"` and `"direct"`:

```{r}
fl_rollup <- cog_geographic_rollup(
  govids = list(
    state  = "120000226351",                   # Florida
    county = c("121011212191", "121099101897") # Broward + Palm Beach
  ),
  category = "Highways",
  years = c(2012, 2020)
)
fl_rollup
```

For the neighboring state, the comparison is a single government, so it's a
plain `cog_spending()` call rather than a rollup:

```{r}
ga_state <- cog_spending(
  "130000226087", years = c(2012, 2020), category = "Highways" # Georgia
)
ga_state
```

Now the same rollup, but asking for `expenditure_concept = "total"`:

```{r, error = TRUE}
cog_geographic_rollup(
  govids = list(state = "120000226351", county = "121011212191"),
  category = "Highways",
  years = 2020,
  expenditure_concept = "total"
)
```

`cog_geographic_rollup()` (and `cog_peer_compare()`, for the same reason)
refuses `"total"` outright rather than silently returning an inflated
number. The next section is why.

# The mechanism

Suppose Alabama gives a county $10M toward a highway project. That $10M
shows up **twice** in the underlying corpus:

- Once on Alabama's own record, coded `M44` ("to local governments,
  Highways") -- Alabama's intergovernmental leg.
- Again on the county's record, coded `E44` / `F44` ("Highways, current
  operations" / "capital outlay") -- the county's direct spending, because
  the county is the government that actually lets the contract and pays the
  paving crew.

`primary` and `direct` (the crosswalk's non-intergovernmental expenditure
subtypes) only ever count the second of those -- the government that
actually did the spending. `total` (Direct plus the intergovernmental leg)
counts the first one *as well*, which is exactly right for describing
Alabama's own budget: Alabama's `total` genuinely includes the $10M it
committed to highways, whether it built the road itself or paid the county
to. But sum `total` across Alabama **and** the county, and that $10M is
counted twice -- once as Alabama's payment out, once as the county's
spending in -- reporting $20M of highway work for $10M actually spent.

This is exactly the shape of query `cog_geographic_rollup()` exists to run
(summing across layers of government), so it refuses `"total"` rather than
silently overstating every multi-layer figure it produces.

# How big is the risk in practice

Intergovernmental transfers aren't evenly distributed by government type.
Measured against the bundled fixture corpus (all 50 states, each of its
four years -- 2011, 2012, 2019, 2020), intergovernmental spending as a
share of a government's own Direct spending is:

| Government type | Intergovernmental / Direct |
|---|---|
| State  | 33.1%-40.5% (varies by year; 36.2% pooled across all four) |
| County | 3.3%-4.8% (varies by year) |
| City   | 2.4%-2.9% (varies by year) |

So the Direct/Total choice matters overwhelmingly for **state**
governments -- a state's Total genuinely differs from its Direct by more
than a third, while for a county or city the two are close. (The state
share is much larger than pre-#11 measurements suggested, because the
intergovernmental leg now correctly includes the `Q11`/`Q12`/`Q18` state
payments to school systems -- for most states the single largest transfer
they make.) That's also why the mistake this vignette warns about is easy
to make unnoticed at the county/city level and costly at the state level:
rolling up every government using `total` instead of `primary`/`direct`
overstates the FY2019 figure by 24.1% for Alabama and 23.2% nationally.

# Why Total = Direct + M + L + Q, not Direct + M

It's tempting to assume `total` only needs to add `M`. But the
intergovernmental leg has three families, all money the queried government
itself pays **out** -- they're not different accounts of a receiving
government's revenue. `M` is what it pays to other **local** governments
(e.g. a county paying a city for a shared paving contract); `L` is what it
pays **up** to its **state** government (e.g. a county's contribution to a
state-administered program); and `Q11`/`Q12`/`Q18` are a state's payments
to **school systems** (K-12 and higher-ed aid -- for most states the
single largest transfer they make, and the piece the pre-#11 prefix
allowlist silently dropped, finding F-017). A government's Total genuinely
includes every leg it pays, because each is its own spending, just routed
to a different kind of recipient. On the bundled fixture corpus (all 50
states, 2011/2012/2019/2020), `L` is 0 for state governments (a state has
no "payments to the state government" leg of its own) but is 43%-51% the
size of `M` for counties (varies by year) and 144%-189% the size of `M`
for cities (varies by year; 166% pooled across all four) -- so a `total`
that omitted `L` would silently undercount Total specifically for local
governments, and for cities `L` is often the *larger* of the two legs.
`cog_spending(expenditure_concept = "total")` includes every leg
(excluding the `L--` family-total rollup row, which would double-count its
own components).

# Composition rules

- `expenditure_concept` (whose spending counts -- Primary, Direct, or
  Direct plus intergovernmental) is **orthogonal** to `basis` (which
  vintage of the item-code space a query resolves against --
  `"harmonized"` vs `"raw"`).
  They combine freely: `expenditure_concept = "total", basis = "raw"` is a
  valid, meaningful query, and so is every other pairing.
- `expenditure_concept = "total"` is **mutually exclusive** with `recipe`: a
  recipe already defines its own component codes (some recipes have their
  own matching intergovernmental counterpart recipe instead -- see
  `cog_recipes()` and the "firing suggestion" notes surfaced in
  `cog_spending()`'s provenance), so layering a second, generic `total`
  union on top of a recipe query has no well-defined meaning. Passing both
  together aborts with an error naming the conflict.
- `expenditure_concept` is a **spending-only** concept: `cog_revenue()`
  doesn't expose it (revenue's own intergovernmental codes are a different
  axis -- see `?cog_revenue`).

# Summary

- Comparing one government to itself over time: any concept works -- pick
  one and hold it fixed across every year compared.
- Comparing or summing across governments -- counties within a state, a
  state against its neighbor, cities against counties: use `"primary"`
  (the default) or `"direct"`. `cog_geographic_rollup()` and
  `cog_peer_compare()` enforce this by refusing `"total"`.
- `"primary"` = operations + capital + assistance. `"direct"` = primary +
  interest on debt + insurance trust benefits (Census's published Direct
  Expenditure). `"total"` = direct + intergovernmental (`M` to local
  governments, `L` to the state government excluding the `L--`
  family-total row, and `Q11`/`Q12`/`Q18` state payments to school
  systems).
